---
title: "Covid-19 Publications"
author: "Tabitha Tatusko"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#import libraries
library(tidyverse)
library(janitor)
library(datasets)
library(tidytext)
library(textdata)
library(stringr)
library(plotly)
library(wordcloud)

#import wrangled data
#pubs_data <- readRDS(file = "data/covid_pubs.RDs")
country_pubs <- readRDS(file = "data/pubs_by_country.RDs")
common_topics <- readRDS(file = "data/common_topics.RDs")

```


During the pandemic, scientists around the globe were putting all of their resources into better understanding Covid-19 and finding a cure or vaccine. This lead to an unprecedented level of scientific advancement and research, which can be reflected in the scientific publications released to document these feats. 

## Publications by Country

Although everyone was working around the clock to find a vaccine, there were some countries that rose above with their amount of contributions.

```{r echo=FALSE}
#ggplot barchart of publications by country
country_pub_gg <- country_pubs %>%
  slice(1:10) %>%
  ggplot(aes(x = fct_reorder(country, count), y = count, color = country, fill = country)) +
  geom_col() +
  facet_grid(. ~ pub_year) +
  # Rotate graph
  coord_flip() +
  guides(color = "none", 
         fill = "none") +
  labs(
    title = "Top Publication Producing Countries in 2020 and 2021",
    x = "Publication Count",
    y = "Country"
  )
 
#interactive plotly chart of publications by country
cpubs_plotly <- ggplotly(country_pub_gg, tooltip = "count") 
cpubs_plotly
```

As you can see, throughout the height of the pandemic the United States was the top contributor of Covid related publications. This data was taken from Dimensions, the world's largest linked research database. 

## Most Common Publication Topics

While the overarching goal was to find a vaccine for Covid, that wasn't necessarily the main focus in all of these publications.

```{r echo=FALSE}
set.seed(12)
#create common topics word cloud
common_topics %>%
  with(wordcloud(words = word, freq = count, max.words = 50))

```

As you can see, many of the most common topics (based on publication title) aren't directly about a vaccine. Many focus on Covid patients and their health, or the severity of infection, or impact of care recieved. These all come together to help us better understand the workings of Covid, which in turn lead to us being able to synthesis a vaccine. 