---
title: "Covid-19 Publications"
author: "Tabitha Tatusko"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#import libraries
library(tidyverse)
library(janitor)
library(datasets)
library(tidytext)
library(textdata)
library(stringr)
library(plotly)

```

```{r include=FALSE}
#initial wrangling of data

#import data
publications_info <- read.csv("data/covid19_publications.csv")

#clean column names
pubs_info <- publications_info %>%
  clean_names()

#wrangle data
pubs_data <- pubs_info %>%
  select(pub_year, country_of_research_organization, title) %>%
  filter(complete.cases(.)) %>%
  filter(!is.na(country_of_research_organization)) %>%
  filter(country_of_research_organization != "")
```


During the pandemic, scientists around the globe were putting all of their resources into better understanding Covid-19 and finding a cure or vaccine. This lead to an unprecedented level of scientific advancement and research, which can be reflected in the scientific publications released to document these feats. 

## Publications by Country

Although everyone was working around the clock to find a vaccine, there were some countries that rose above with their amount of contributions.

```{r echo=FALSE}
#create publication prevalence by country barchart
country_pubs <- pubs_data %>%
  rename(country = country_of_research_organization) %>%
  unnest_tokens(output = country, input = country, 
                token = 'regex', pattern=";", to_lower = F) %>%
  mutate(across(where(is.character), str_trim)) %>%
  #this removes duplicate countries for one publication
  #ex: first title had 5 Kuwait authors, but still only 1 pub coming from Kuwait
  group_by(country) %>%
  filter(title!=lag(title) | is.na(lag(title))) %>%
  group_by(pub_year) %>%
  count(country, sort = T) %>%
  rename(count = n)

#ggplot barchart of publications by country
country_pub_gg <- country_pubs %>%
  slice(1:10) %>%
  ggplot(aes(x = fct_reorder(country, count), y = count, color = country, fill = country)) +
  geom_col() +
  facet_grid(. ~ pub_year) +
  # Rotate graph
  coord_flip() +
  guides(color = "none", 
         fill = "none") +
  labs(
    title = "Top Publication Producing Countries in 2020 and 2021",
    x = "Publication Count",
    y = "Country"
  )
 
#interactive plotly chart of publications by country
cpubs_plotly <- ggplotly(country_pub_gg, tooltip = "count") 
cpubs_plotly
```

As you can see, throughout the height of the pandemic the United States was the top contributor of Covid related publications. This data was taken from Dimensions, the world's largest linked research database. 

## Most Common Publication Topics

While the overarching goal was to find a vaccine for Covid, that wasn't necessarily the main focus in all of these publications.

```{r echo=FALSE}
#create top words barchart
data(stop_words)

#seperate words from titles
common_topics_all <- pubs_data %>%
  select(!country_of_research_organization) %>%
  unnest_tokens(output = word, input = title)

#remove common stop words and additional covid related stop words
common_topics <- common_topics_all %>%
  anti_join(stop_words, by = "word") %>%
  count(word, sort = T) %>% #also removed spanish stop words de/la
  filter(!word %in% c("covid" , "19", "pandemic", "sars", "cov", "2", "coronavirus", "2020", "2019",
                      "based", "disease", "de", "la")) %>% 
  rename(count = n)

#create bar chart of top 15 most common words in titles
topics_gg <- common_topics %>%
  slice(1:10) %>%
  ggplot(aes(x = fct_reorder(word, count), y = count, color = word, fill = word)) +
  geom_col() +
  # Rotate graph
  coord_flip() +
  guides(color = "none", 
         fill = "none") +
  labs(
    title = "Common Topics in Covid Publications",
    y = "Number of Occurences",
    x = "Topic Word"
  )

#ggplot to plotly
topics_plotly <- ggplotly(topics_gg, tooltip = "count")
topics_plotly
```

As you can see, many of the most common topics (based on publication title) aren't directly about a vaccine. Many focus on Covid patients and their health, or the severity of infection, or impact of care recieved. These all come together to help us better understand the workings of Covid, which in turn lead to us being able to synthesis a vaccine. 