---
title: "Worldwide analysis of covid-19 related factors and the socio-economic determinants
  of countries"
author: "Lynca Jones Kaminka"
date: "\today"
output:
  html_document:
    df_print: paged
  pdf_document: default
subtitle: 'SAT 231: Blog Project'
---

```{r setup, include = FALSE}
# set code chunk option defaults
knitr::opts_chunk$set(
  # display code as types
  tidy = FALSE, 
  # slightly smaller code font
  size = "small",
  # do not display messages in PDF
  message = FALSE,
  # set default figure width and height
  fig.width = 5, fig.height = 3) 

# improve digit and NA display 
options(scipen = 1, knitr.kable.NA = '')

# load packages
library(dplyr)
library(tidyverse)
library(tidyr)
library(ical)
library(lubridate)
library(kableExtra)
library(dplyr) #for inner join, renaming columns, #rounding
library(DT)
library(ggrepel)
library(tidytuesdayR)
library(plotly)#hovering
library(shinythemes)
library(stringi)
library(ggthemes)
library(gridExtra) ##beautifying the graph ~ rotating the labels.
library(knitr)
library(ggplot2)
library(RMariaDB)
library(lubridate)

# help with extracting the right data frames from xlsx
library(readr)
library(tidyverse)
library(sf)

library(rvest) #for web scrapijng
library(robotstxt) #checking if paths are allowed
library(purrr)

#centroids. 

library(rgeos)
library(rworldmap)
library(rworldxtra)

library(crosstalk) #for crosstalk. 
library(htmlwidgets)
library(bslib) # framework that allows one to styel theire Shiny applications and R Markdown documents with pre-build Bootstrap (open-source library created by Twitter) themes.

library(reactable)
library(leaflet)
```


```{r, echo = FALSE}

#first data frame
covid_worldwide <- read_csv("data/owid-covid-data.csv")


world_map <- maps::map("world", plot = FALSE, fill = TRUE) %>%
st_as_sf()


# get world map
wmap <- getMap(resolution="high")

# get centroids
centroids_df <- gCentroid(wmap, byid=TRUE)


#determining where the mismatch happens: 
covid_worldwide_mismatch_1 <- covid_worldwide %>%
  anti_join(world_map, by=c("location" = "ID"))


covid_worldwide_mismatch_2 <- world_map %>%
  anti_join(covid_worldwide, by=c("ID" = "location"))

#resolving the mismatch by changing the country names. 
covid_worldwide_1 <- covid_worldwide %>%
mutate(location = as.character(location)
, location = case_when(location == "Democratic Republic of Congo" ~ "Democratic Republic of the Congo"
, location == "Congo" ~ "Republic of Congo"                       
,  location == "Eswatini" ~ "Swaziland" 
, location == "Cote d'Ivoire" ~ "Ivory Coast"
, location == "Slovak Republic" ~ "Slovakia"
, location == "United States" ~ "USA"
, location == "United Kingdom" ~ "UK"
, location == "United States Virgin Islands" ~ "Virgin Islands, US"
, location == "Trinidad and Tobago" ~ "Tobago"
#, country == "Trinidad and Tobago" ~ "Trinidad"
, location == "Myanmar" ~ "Myanmar"
, TRUE ~ location))



#joining the two data sets: 
covid_worldwide_2 <- covid_worldwide_1 %>%
  right_join(world_map, by=c("location" = "ID"))


covid_worldwide_3 <- covid_worldwide_2 %>%
separate("date", into = c("ReportYear", "ReportMonth", "ReportDay"), sep = "-", convert=FALSE) %>%
mutate(ReportYear = as.numeric(ReportYear))%>%  
mutate(ReportDay = as.numeric(ReportDay))%>%  
mutate(ReportMonth = as.numeric(ReportMonth))

covid_worldwide_4 <- covid_worldwide_3 %>%
filter(ReportYear%in%c(2021,2022))
    
    
covid_worldwide_5 <- covid_worldwide_4 %>%
mutate(date = make_date(year = covid_worldwide_4$ReportYear,
                month = covid_worldwide_4$ReportMonth,
                day = covid_worldwide_4$ReportDay))

    # helped me with figuring out which observations to keep and which ones to get rid of 
    #colnames(covid_worldwide_4)
    
#type_data1: identifying factors
#data2: covid_related factors
#data3:    socio-economic factors
  #names(covid_worldwide_5)
   
covid_worldwide_6 <- covid_worldwide_5 %>%
select(iso_code | "location"| "continent" | "date" | geom |
      "total_cases_per_million" | "total_cases" |
      "people_vaccinated" | "new_deaths_per_million"  |"new_cases_smoothed" | "new_cases_per_million"  
      | "new_deaths_per_million" | "new_deaths_smoothed" |     "new_tests_smoothed"| "new_vaccinations_smoothed" |
      "population_density"  | "life_expectancy" | "gdp_per_capita" | "aged_65_older"
      | "population" 
      | "extreme_poverty" | "diabetes_prevalence")  

    
#extracting the latitudes and longitudes observations for each country .
 # 1. Identify where the table is
countries_worldwide_lat_long_url <- "https://developers.google.com/public-data/docs/canonical/countries_csv"

# 2. Confirm bots are allowed to access the page 
robotstxt::paths_allowed(countries_worldwide_lat_long_url)


countries_worldwide_lat_long <- countries_worldwide_lat_long_url %>%
read_html() %>%
# Poems are in first table on page
html_elements("table") %>%
purrr::pluck(1) %>%
html_table()
 

#standardizing countries'names in the new data set. 
countries_worldwide_lat_long_updated <- countries_worldwide_lat_long %>%
mutate(country = as.character(name))

  
#checking whether some country names of the two data tables that we want to join are mismatched?
covid_worldwide_mismatch_3 <-  covid_worldwide_6%>%
  anti_join(countries_worldwide_lat_long_updated, by=c("location" = "country"))
#unique(covid_worldwide_mismatch_3$location)

covid_worldwide_mismatch_5 <- countries_worldwide_lat_long_updated %>%
  anti_join(covid_worldwide_6, by=c("country" = "location"))
#unique(covid_worldwide_mismatch_5$name) 

 covid_worldwide_6 <- covid_worldwide_6 %>%
mutate(location = as.character(location)
, location = case_when(location == "Democratic Republic of the Congo" ~ "Congo [DRC]"
, location == "Republic of Congo" ~ "Congo [Republic]"                        
, location == "Ivory Coast" ~ "Côte d'Ivoire"
, location == "USA" ~ "United States"
, location == "UK" ~ "United Kingdom"
, location ==  "Virgin Islands, US" ~ "U.S. Virgin Islands"
# location == "Trinidad and Tobago" ~ "Tobago"
#, location == "Trinidad and Tobago" ~ "Trinidad"
, location == "Palestine" ~ "Palestinian Territories"
, location ==  "Myanmar" ~ "Myanmar [Burma]"
, TRUE ~ location))

 #joining the two data frames, one containing the covid-19 related observations and the other information about the longitude and the latitude of each polygon/country.  
 covid_worldwide_7 <- countries_worldwide_lat_long_updated %>%
 right_join(covid_worldwide_6, by=c("country" = "location"))
 
 covid_worldwide_7_missing_data_12 <- covid_worldwide_7 %>%
   select(country, continent, new_cases_smoothed, new_deaths_smoothed, new_tests_smoothed)
 
#identifying the columns with numeric types, and replacing the observation from na to zero in that case. 
 covid_worldwide_missing_data <- covid_worldwide_7 %>%
   group_by(country, continent, gdp_per_capita)%>%
   summarize(missing_new_cases = sum(is.na(new_cases_smoothed)), 
         missing_new_deaths_smoothed = sum(is.na(new_deaths_smoothed)), 
         missing_new_tests_smoothed = sum(is.na(new_deaths_smoothed)))

covid_worldwide_8 <- covid_worldwide_7 %>%
  mutate_if(is.numeric, ~ifelse(is.na(.), 0, .))

saveRDS(covid_worldwide_missing_data, file = "data/covid_worldwide_missing_data.RDS")
saveRDS(covid_worldwide_8 , file = "data/covid_worldwide_data.Rds")

```




## Analysis of socio-economic factors that have influenced covid-19 cases from a global perspective.

In my analysis, I delved into the distribution of covid-19 metrics, such as the number of deaths, cases and vaccinations, and took a country-by-country approach to comparing these values. Through this process, I created a graph that visualizes the total number of covid cases reported by all countries from January 2020 to January 2023. Additionally, the accompanying table provides a more detailed overview of pandemic observations, including the number of cases, deaths and vaccinations adjusted to the population size, allowing for a closer examination of which countries have fared better or worse. 

The goal of creating this graph was to first recognize the disparity in reported per capita COVID-19 cases across different countries. This lays the foundation for further analysis of the potential socio-economic factors that may contribute to these trends.


```{r, warnings=FALSE, message=FALSE, echo = FALSE}
options(warn=-1)
covid_worldwide_9 <- covid_worldwide_8 %>%
  filter(!is.na(country) & country != "")

#total counts
  covid_worldwide_dataT <- covid_worldwide9 %>%
    select("country" | "continent" | "date"| "new_deaths_smoothed" |
             "new_cases_smoothed" | "new_tests_smoothed"| "new_vaccinations_smoothed" | "population" |"life_expectancy"|"aged_65_older" | iso_code | geom | longitude | latitude) %>%
    group_by(country, continent, population, life_expectancy, aged_65_older, latitude, longitude, geom, iso_code)%>%
    summarize(total_cases = sum(new_cases_smoothed), 
              total_deaths = sum(new_deaths_smoothed), 
              total_tests = sum(new_tests_smoothed), 
              total_vaccinations = sum(new_vaccinations_smoothed))%>%
    ungroup()
  
    
  #per capita counts
    covid_worldwide_dataT_capita <- covid_worldwide_dataT %>%
    select("country" | "continent" | "total_cases" |
             "total_deaths" | "total_tests"| "total_vaccinations" | "population" |"life_expectancy"|"aged_65_older" | iso_code | geom | longitude | latitude) %>%
    group_by(country, continent, population, life_expectancy, aged_65_older, latitude, longitude, geom, iso_code)%>%
    summarize(total_cases_capita = round((total_cases/population) * 100, 3),
              total_deaths_capita = round((total_deaths/population) *100, 3),
              total_tests_capita = round((total_tests/population) *100, 3),
              total_vaccinations_capita = round((total_vaccinations/population)*100,3))%>%
      ungroup()
    
    
#extracting only the desired pieces of information from the data set.
 covid_worldwide_leaflet_map <- covid_worldwide_dataT_capita %>%
   select("country", "continent", longitude, latitude, geom, total_cases_capita, total_deaths_capita)
 

 #transforming it into a leaflet object: 
 covid_worldwide_7_leaflet <- covid_worldwide_leaflet_map %>%
   st_as_sf()
 

 #caused me so much trouble but it is use to transform spatial data from one coordinate reference system (CRS) to another.From CRS to WGS84
 covid_worldwide_8_leaflet <- st_transform(covid_worldwide_7_leaflet, "+proj=longlat +datum=WGS84")
 

# Define a color palette 
my_palette <- colorQuantile(rev(viridis::viridis(20))
                  , domain =  covid_worldwide_8_leaflet$total_cases_capita, n=20)

# Create interactive map, zoomed in on Amherst
map <- leaflet(data = covid_worldwide_8_leaflet
         , options = leafletOptions(attributionControl = FALSE)) %>%
 # addTiles() %>%
   addPolygons(
      fillColor = ~my_palette(total_cases_capita),
       stroke = FALSE,
        popup = ~ glue::glue("Country: {str_to_title(country)}
                        <br> {paste0(round(total_cases_capita,2), '%')}")
      , weight = 2, smoothFactor = 0.5, opacity = 1, fillOpacity = 1
      , labelOptions= labelOptions(direction = 'auto')
      , color='#333333'
      , highlightOptions = highlightOptions(color='#000000', weight = 2
                                            , bringToFront = TRUE
                                            , sendToBack = TRUE)) %>%
   addLegend(pal = my_palette,
            values = covid_worldwide_8_leaflet$total_cases_capita,
           position = "topright",
           title = "Percentage of Per capita cases reported
            <br> (% 2021-2022)", opacity = 1)

```

```{r, warnings=FALSE, message=FALSE, echo = FALSE}
options(warn=-1)
# ungroup your dataframe
covid_worldwide_dataT_capita_ungrouped <- covid_worldwide_dataT_capita %>% 
  ungroup()
covid_worldwide_dataT_capita_table <- covid_worldwide_dataT_capita_ungrouped %>% 
  select("country" , total_cases_capita , total_deaths_capita , total_vaccinations_capita , total_tests_capita) 

#covid_worldwide_dataT_capita_table%>% 
  dataframe <- DT::datatable(covid_worldwide_dataT_capita_table, 
                colnames = c("Country", "Percentage of Covid-19 cases Per Capita", 
                         "Percentage of  Deaths Per Capita", 
                         "Percentage of  Vaccinations Per Capita", 
                         "Percentage of  Tests Per Capita"),
            filter = 'top',
            options = list(autoWidth = TRUE))

```



#### Country-level covid-19 view {.tabset}

##### Spatial View

```{r, warnings=FALSE, message=FALSE, echo = FALSE}
map
```
##### Table View

```{r, warnings=FALSE, message=FALSE, echo = FALSE}
dataframe
```

#### Some features about both representations: 

At a glance, the majority of countries are reporting that the total number of people who have caught covid is
between 10 and 20% of their national population. However the table is worrying as some countries claim to have never had a covid case like North Korea or never had a covid death like my home country, Burundi, but I was there last summer and I can assure you that far too many people had "the flu". Some countries claim to have never done a test or vaccinations. 

The other notable feature of the map is that some countries in Europe such as France, Portugal, Iceland, South Korea, Bulgaria, and others, appear to have had the highest cumulative per capita COVID-19 cases from 2021 to 2022.

#### Covid distribution in relation to other factors. 

As my first objective was to find a potential connecting thread between the socio-economic factors specific to each country and the fact that they were reflected in the observations associated with the pandemic, I undertook to analyze the progression of covid- 19 for each country and I have included filtering choices which allowed to visualize the countries whose socio-economic factor is in a certain range, say whose GDP is in a certain range, average age and so on. 


#### Socio-economic factors and covid related observations.

##### Per Capita Percentage of Case Counts vs GDP Per Capita and Population Density

```{r, warnings=FALSE, message=FALSE, echo = FALSE}
covid_worldwide_8 <- covid_worldwide_7 %>%
  filter(!is.na(country) & country != "")


covid_worldwide_9 <- covid_worldwide_8 %>%
  group_by(country, continent, date, extreme_poverty, gdp_per_capita, aged_65_older, life_expectancy,           
           diabetes_prevalence, population_density, population)%>%
    summarize(new_cases_smoothed_capita = round((new_cases_smoothed/population) *100, 3),
              new_deaths_smoothed_capita = round((new_deaths_smoothed/population) *100,3), 
              new_tests_smoothed_capita = round((new_tests_smoothed/population)*100,3),
              new_vaccinations_smoothed_capita = round((new_vaccinations_smoothed/population)*100,3))%>%
  ungroup()

covid_lines <- highlight_key(covid_worldwide_9)

gg_covid <- ggplot (data = covid_lines, aes(x=date, y = new_cases_smoothed_capita, group = country)) +
  geom_line(aes(color = continent,            
                text = paste("Country: ", country, 
                             "<br>Date: ", date, 
                              "<br> New covid cases per capita", new_cases_smoothed_capita, "%",
                              "<br> Population Density", population_density, 
                              " <br> GDP per capita", gdp_per_capita
                        )), show.legend = FALSE)+
  xlab("Date") + ylab("Percentage of New Cases Per Capita")+ 
  ggtitle("Per Capita Covid Cases Trend") +
  theme_bw()

 bscols(widths = c(3,9),
        list(
              filter_select("id", label = "Select a country", covid_lines, ~country),
              filter_slider(id = "slider_pd", label = "Population Density",
                    covid_lines, column = ~population_density),
              filter_slider(id = "slider_gdp_per_capita", label = " GDP Per Capita", 
                            covid_lines, column = ~gdp_per_capita)
             ), 
          ggplotly(gg_covid, dynamicTicks = TRUE, tooltip = c("text"),  
           tooltipoptions = list(placement = "bottom"))
 
)

```


##### Per Capita Percentage of  Covid-induced Deaths vs GDP Life Expectancy and Proportion of 65+ people.

```{r, warnings=FALSE, message=FALSE, echo = FALSE}
covid_lines_2 <- highlight_key(covid_worldwide_9)

gg_covid_2 <- ggplot (data = covid_lines_2, aes(x=date, y = new_deaths_smoothed_capita, group = country)) +
  geom_line(aes(color = continent,            
                text = paste("Country: ", country, 
                             "<br>Date: ", date, 
                              "<br> New covid cases per capita", new_deaths_smoothed_capita, "%", 
                              "<br> Life expectancy", life_expectancy, 
                             " <br> Aged 65 older", aged_65_older
                              
                              
                        )), show.legend = FALSE)+
  xlab("Date") + ylab("Percentage of New Deaths Per Capita")+ 
  ggtitle("Per Capita Covid Deaths Trend") +
  theme_bw()

 bscols(widths = c(3,9),
        list(
              filter_slider(id = "slider_le", label = "Life Expectancy", 
                            covid_lines_2, column = ~life_expectancy),
              filter_slider(id = "slider_aged70", label = "Aged 65 older", 
                            covid_lines_2, column = ~aged_65_older)
             ), 
          ggplotly(gg_covid_2, dynamicTicks = TRUE, tooltip = c("text"),  
           tooltipoptions = list(placement = "bottom"))
 
)

```
##### Per Capita Percentage of Vaccination Counts vs GDP Per Capita for each country. 

```{r}
covid_lines_3 <- highlight_key(covid_worldwide_9)

gg_covid_3 <- ggplot (data = covid_lines_3, aes(x=date, y = new_vaccinations_smoothed_capita, group = country)) +
  geom_line(aes(color = continent,            
                text = paste("Country: ", country, 
                             "<br>Date: ", date, 
                              "<br> Vaccinations per capita", new_vaccinations_smoothed_capita, "%",
                              " <br> GDP per capita", gdp_per_capita
                        )), show.legend = FALSE)+
  xlab("Date") + ylab("Percentage of New Vaccinations Per Capita")+ 
  ggtitle("Per Capita Vaccinations Trend") +
  theme_bw()

 bscols(widths = c(3,9),
        list(
              filter_select("id", label = "Select a country", covid_lines_3, ~country),
              filter_slider(id = "slider_gdp_per_capita_2", label = " GDP Per Capita", 
                            covid_lines_3, column = ~gdp_per_capita)
             ), 
          ggplotly(gg_covid_3, dynamicTicks = TRUE, tooltip = c("text"),  
           tooltipoptions = list(placement = "bottom")))
```

#### Note: 
The pay to price of an interactivity that is not embedded within shiny is that the interactivity is very slow, it doesn't allow for a smooth use of the graph. .

#### Possible Findings: 

#### Tools and Inherent Limitations associated with the tools.
Crosstalk, according to Chapter 16 of https://plotly-r.com/client-side-linking.html  is a framework for creating widgests and dashboards that allow users to filter, sort and group data, and then visualize the results in real-time.Consequently, it allows one to embed an interactive histogram into an html page generated from an R Markdown document, which is what I needed. 

Due to its computational demands, crosstalk isn't be suitable for larger datasets with a high number of daily observations. Thus, my analysis of the trend of different COVID observations was limited.

#### Inherent Limitation to Making Conclusive Findings: 
During the data wrangling process, it became apparent that many countries had no COVID-related observations for several months at a time. This lack of reported differs from a zero observation, reason why I haven't made any conclusion yet, and I believe that it should be taken into account when analyzing the data.



```{r}
#thanks to: for controlling -tooltips: https://plotly-r.com/controlling-tooltips.html and for the introduction to #cross-talk:  https://plotly-r.com/client-side-linking.html 
```


















