---
title: "emmawranglecovid"
author: "Emma Strawbridge"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(tidyverse)
library(dplyr)
library(readxl)

bigBoy <- read_excel("~/Downloads/covid-19-raw-data-2-16-2023.xlsx", sheet = 4)
poverT <- read.csv("~/data/covid-19-raw-data-2-16-2023.xlsx", sheet = 10)
```

## things to note

Definition of Poverty rate: https://data.oecd.org/inequality/poverty-rate.htm#indicator-chart 
The poverty rate is the ratio of the number of people (in a given age group) whose income falls below the poverty line; taken as half the median household income of the total population. It is also available by broad age group: child poverty (0-17 years old), working-age poverty and elderly poverty (66 year-olds or more). However, two countries with the same poverty rates may differ in terms of the relative income-level of the poor.

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
